OTCE: Hybrid SSM and Attention with Cross Domain Mixture of Experts to construct Observer-Thinker-Conceiver-Expresser

### 2025-01-12 paper.tex
icaml_

\begin{icmlauthorlist}
\icmlauthor{Firstname1 Lastname1}{equal,yyy}
\icmlauthor{Firstname2 Lastname2}{equal,yyy,comp}
\icmlauthor{Firstname3 Lastname3}{comp}
\icmlauthor{Firstname4 Lastname4}{sch}
\icmlauthor{Firstname5 Lastname5}{yyy}
\icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
\icmlauthor{Firstname7 Lastname7}{comp}
%\icmlauthor{}{sch}
\icmlauthor{Firstname8 Lastname8}{sch}
\icmlauthor{Firstname8 Lastname8}{yyy,comp}
%\icmlauthor{}{sch}
%\icmlauthor{}{sch}
\end{icmlauthorlist}

@@

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except that
author information (names and affiliations) should be given. See
\cref{final author} for formatting instructions.

The footnote, ``Preliminary work. Under review by the International
Conference on Machine Learning (ICML). Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{42}^{nd}$ International Conference on Machine Learning},
Vancouver, Canada, PMLR 267, 2025.
Copyright 2025 by the author(s).''

For those using the \textbf{\LaTeX} style file, this change (and others) is
handled automatically by simply changing
$\mathtt{\backslash usepackage\{icml2025\}}$ to
$$\mathtt{\backslash usepackage[accepted]\{icml2025\}}$$
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one. The running title consists of a
single line centered above a horizontal rule which is $1$~point thick.
The running head should be centered, bold and in $9$~point type. The
rule should be $10$~points above the main text. For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the \texttt{fancyhdr} package which is included in the ICML
2025 style file package. In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the specified format.

\subsection{Dimensions}

The text of the paper should be formatted in two columns, with an
overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
between the columns. The left margin should be 0.75~inches and the top
margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.
Do not write anything on the margins.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

ICML uses double-blind review, so author information must not appear. If
you are using \LaTeX\/ and the \texttt{icml2025.sty} file, use
\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
will not be printed unless \texttt{accepted} is passed as an argument to the
style file.
Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If you are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we
have shown \ldots'').

Do not anonymize citations in the reference section. The only exception are manuscripts that are
not yet published (e.g., under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have
to be submitted
as Supplementary Material via OpenReview\@. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look at the Supplementary Material when writing their
review (they are not required to look at more than the first $8$ pages of the submitted document).

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches below the
bottom rule surrounding the title. The authors' names should appear in 10~point
bold type, in a row, separated by white space, and centered. Author names should
not be broken across lines. Unbolded superscripted numbers, starting 1, should
be used to refer to affiliations.

Affiliations should be numbered in the order of appearance. A single footnote
block of text should be used to list all the affiliations. (Academic
affiliations should list Department, University, City, State/Region, Country.
Similarly for industrial affiliations.)

Each distinct affiliations should be listed once. If an author has multiple
affiliations, multiple superscripts should be placed after the name, separated
by thin spaces. If the authors would like to highlight equal contribution by
multiple first authors, those authors should have an asterisk placed after their
name in superscript, and the term ``\textsuperscript{*}Equal contribution"
should be placed in the footnote block ahead of the list of affiliations. A
list of corresponding authors and their emails (in the format Full Name
\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
Ideally only one or two names should be listed.

A sample file with author names is included in the ICML2025 style file
package. Turn on the \texttt{[accepted]} option to the stylefile to
see the names rendered. All of the guidelines above are implemented
by the \LaTeX\ style file.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below the final
address. The heading `Abstract' should be centered, bold, and in 11~point type.
The abstract body should use 10~point type, with a vertical spacing of
11~points, and should be indented 0.25~inches more than normal on left-hand and
right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
sentences. Gross violations will require correction at the camera-ready phase.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}
Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{Footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper.
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
\caption{Historical locations and number of accepted papers for International
Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
produced, the number of accepted papers for ICML 2008 was unknown and instead
estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to illustrate
your approach and results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
\emph{after} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
\cref{icml-historical}. The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left. You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment \texttt{figure*} in \LaTeX). Always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
environments to format pseudocode. These require
the corresponding stylefiles, algorithm.sty and
algorithmic.sty, which are supplied with this package.
\cref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$}
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like
figures, these should be centered, legible, and numbered consecutively.
However, place the title \emph{above} the table with at least
0.1~inches of space before the title and the same after it, as in
\cref{sample-table}. The table title should be set in 9~point
type and centered unless it runs two or more lines, in which case it
should be flush left.

@@

\bibliography{example_paper}
\bibliographystyle{icml2025}

>>>>

\begin{icmlauthorlist}
\icmlauthor{Jingze Shi}{equal,yyy}
\icmlauthor{Bingheng Wu}{equal,yyy,comp}
\icmlauthor{Ting Xie}{comp}
\icmlauthor{Chunjun Zheng}{sch}

%\icmlauthor{}{sch}
%\icmlauthor{}{sch}
\end{icmlauthorlist}

@@

\subsection{Positional Information}

We first identified a key challenge in combining the Selective State Model with self-attention: the effective integration of positional information. In Mamba, positional information is provided by the implicit local positional information from the causal convolution, while self-attention itself cannot provide positional information, it relies on positional encoding to provide global positional context. To address this issue, we designed a relative positional information injection method that connects the inner product state of the Selective State Space with the inner product state of self-attention, allowing the input gate-state-output gate of the Selective State Space to make filtered relevant information have discrete relative positional information, and in self-attention, the discrete relative positional information is re-continuous to build long-term dependency relationships of relevant information. This method not only enables our model to have the ability to selectively process input sequences but also effectively capture long-distance dependency relationships. In complex multi-query associative recall tasks, our model trained on the same dataset outperforms larger-scale Mamba, Transformer models, and models that mix Mamba and Transformer without our proposed relative positional information injection method.

\subsection{Cross-Domain Mixture of Experts}
In human society, knowledge is widely distributed across different domains, and these domains are interconnected through common foundational knowledge and cross-domain connections. To simulate this phenomenon in the model, we designed two types of cross-domain mixed experts: Cohesive Cross-Domain Expert and Expansive Cross-Domain Expert. These experts (multi-layer perceptrons) store and transfer common foundational knowledge and cross-domain knowledge between different domains by sharing parameters. The Cohesive Cross-Domain Expert achieves close integration between domains by sharing linear layer parameters within all experts, which is more suitable for small-scale models with fewer experts because of its faster computation speed. The Expansive Cross-Domain Expert shares a complete MLP parameter, adding a common domain knowledge gate in each expert to control the flow of common domain knowledge into private MLP parameters, which is more suitable for large-scale models with more experts because it allows for more flexible adjustment and utilization of common knowledge. Experimental results show that the performance of these two cross-domain mixed experts on the same dataset is better than shared expert isolated mixed experts, confirming the effectiveness of our design in promoting cross-domain knowledge transfer and improving model generalization.

\subsection{Architecture Design}

From a biological perspective, the relationship between input and output can be described by observing, thinking, conceiving, and expressing four stages. Inspired by this idea, we designed a new architecture with a bionic perspective: Observer-Thinker-Conceiver-Expresser (\textbf{OTCE}). The OTCE architecture mimics the natural process of information processing in biology, aiming to optimize information processing and transmission in a modular way. In the Observer module, we use the selective state space's selection ability to filter out irrelevant information in the sequence to retain relevant information. In the Thinker module, we use the self-attention's ability to capture dependencies between any two elements in a sequence of any length, regardless of their position in the sequence, to build long-term dependency relationships. In the Conceiver module, we use the linear recursive state update mechanism of the state space to build a single summary state information. In the Expresser module, we combine the context-aware state information produced by self-attention considering all elements with the summary state information to build a context-aware summary state. We also studied the combination of the Selective State Space and self-attention with ordinary multi-layer perceptrons and cross-domain mixed experts, and finally determined a model combination with the lowest perplexity at the same parameter scale.

We empirically validated OTCE on multiple tasks, including semantic similarity evaluation, long-short text classification, natural language inference, keyword recognition, different domain selection tasks, context learning, and multi-query associative recall tasks. These experiments demonstrate the effectiveness of the OTCE architecture in handling complex language tasks.

\begin{figure}[H]
   \centering
   \includegraphics[width=1.0\linewidth]{fig/otce.png}
   \caption{(\textbf{OTCE Architecture}.) OTCE demonstrates the overall combined architecture and process of using Observer, Thinker, Conceiver, and Expresser modules in language modeling tasks. Observer, Thinker, Conceiver, and Expresser show their internal combination of selective state space, self-attention, multi-layer perceptron, and cross-domain mixed experts.}
   \label{fig:otce}
\end{figure}

\section{Background}
\subsection{Selective State Space Models}

Selective state space models~\cite{gu2023mamba} consider that a fundamental problem in sequence modeling is to compress the context into a smaller state. From this perspective, the attention mechanism of Transformers explicitly stores the entire context information, as if reviewing all previous inputs and generated tokens before writing each token. In contrast, RNNs only refer to a fixed number of previous tokens each time, which allows for faster writing but may forget key tokens.

In the precursor state space models of selective state space models, there are 4 parameters$(\Delta t, A, B, C)$, and they do not change with the input. These parameters control the following two stages:

@@

\bibliography{paper}
\bibliographystyle{icml2025}
